---
title: Plotting with R Base Graphics
teaching: ?
exercises: ?
questions:
- "How can I visualize my data with R?"
objectives:
- survey high-level plotting commands
- review common options for plotting
- using subsetting and selection to add information to plots
- low-level plotting for fine-tuning and clarity
- interacting with your plots
- making plots pretty
- generating output for different purposes
keypoints:
- "R plotting links data to display - if it's in your data, we can plot it!"
- "Plotting isn't special - the code we use for analysis works the same here."
---

```{r chunk_options, include=FALSE}
source("bin/chunk-options.R")
knitr_fig_path("02-")
setwd("~/teaching/R-ecology-lesson/tyler-pipeline/")
```

## Basic plotting

We'll start slow, and build our way up. For starters, consider a short vector of
numbers: 

```{r }
myvect <- 10:1                          # What does 10:1 do?
```

```{r Plotting vectors}
plot(myvect)
```

R guesses what kind of plot you want based on the class of the object you give
it:

```{r}
class(myvect)
```

By default, the values of one-dimensional vectors are plotted in order, almost
like time-series.

```{r Plotting data.frames}
class(women)

plot(women)
```

2-D `data.frames` and matrices with numerical data are plotted as
scatter plots by default.

> ## Quick plots:
>
> R provides default plots for most types of objects. Try `plot(object)` to see
> this default. It's probably not something you'd publish, but it's often a
> useful summary of your data. The `plot` command doesn't change your data, so
> it's safe to experiment as much as you like.

> ## Challenge:
>
> Look at the built-in `iris` data set. Don't plot it yet - first take a look at
> the data. What class are each of the columns? How would you summarize this?
> Find out what the default plot looks like!

## Scatterplots

Let's continue with the `iris` data. We've seen the default plot, but now we
want to take a closer look at `Sepal.Length` and `Sepal.Width`. To get that, we
need to be more explicit with our instructions:

```{r Iris Sepal Scatterplot}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width)
```

```{r ggplotting vectors}
library(ggplot2)
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
```

Here we've specified which variable is goes on each axis. If we leave that out,
R will assume the first variable is `x` and the second is `y`.

Because plotting two variables out of a `data.frame` is such a common request, R
provides some handy short-cuts for doing this:

```{r Scatterplot Shortcuts}
plot(Sepal.Width ~ Sepal.Length, data = iris)
```

The tilde symbol, `~`, is what R uses to indicate the formula of a linear model.
In this case, we're saying "plot `Sepal.Width` **against** `Sepal.Length`". For
scatterplots, that means the variable before the `tilde` is plotted on the
y-axis, and the variable after the tilde is plotted on the x-axis. We don't need
to specify that these are columns of the `iris` data.frame (i.e.,
`iris$Sepal.Length`), because we use the `data` argument to tell R where to
look for the formula variables.

The following commands all do the same thing:

```{r Scatterplot Alternatives}
plot(iris$Sepal.Length, iris$Sepal.Width)
plot(y = iris$Sepal.Width, x = iris$Sepal.Length)
plot(iris[, "Sepal.Length"], iris[,"Sepal.Width"])
plot(iris[, 1:2])
```

> ## Challenge:
>
> The first three variations are just different ways of selecting the same two
> columns for `x` and `y`. What does the fourth line mean in English?

### Colours, Symbols and Size

Now we have a boring scatterplot to display two of our variables. We can modify
the colours, symbols, and size of our points with the arguments `col` (for
colour), `pch` (for plot character), and `cex` (for character expansion). Let's
start with `pch`.

There are 25 different symbols for plotting in R:

```{r pch}
plot(1:25, pch = 1:25)
```

If we set pch to a single value, it will apply to all points in the figure:

```{r one pch}
plot(Sepal.Width ~ Sepal.Length, data = iris, pch = 19)
```

That looks better, but it doesn't add any information to the plot. If we use a
vector of numbers to set `pch`, each point can have a different shape:

```{r many pch}
plot(Sepal.Width ~ Sepal.Length, data = iris, pch = 1:25)
```

This plot uses a different symbol for each of the first 25 points, then starts
over. Again, it doesn't add any value to the plot.

```{r pch by factor}
plot(Sepal.Width ~ Sepal.Length, data = iris, pch = as.numeric(Species))
```

```{r ggplot pch}
ggplot(data = iris,
       aes(x = Sepal.Length, y = Sepal.Width, color = Species,
           shape = Species, size = Petal.Length)) +
  geom_point(alpha = 0.5) +
  geom_abline(aes(intercept = 0, slope = 1))
```

Now the symbols mean somthing useful! Note that this uses the first three
symbols that R knows. If we don't like these, we can do a little selection trick
to pick our own:

```{r pch pick symbols}
plot(Sepal.Width ~ Sepal.Length, data = iris,
     pch = c(19, 24, 4)[as.numeric(Species)])
```

In addition to the 25 symbols, we can also use any single character:

```{r pch letters}
plot(Sepal.Width ~ Sepal.Length, data = iris,
     pch = c("S", "E", "I")[as.numeric(Species)])
```

> ## Challenge
>
> The `col` argument works much the same as `pch`. Find out how many colours are
> available, and set different colours for each species in our plot.

There are actually many more colours available in R than what you can get using
numbers. You can also use names, the full list is displayed by the function
`colors()`.

Another useful tool, particularly when you have large data sets, is
transparency. This is allows us to see when points are overlapping. The easiest
way to get transparency is by 'adding' it to existing colors with the function
`adjustcolor`:

```{r adjustcolor}
adjustcolor(1:4, alpha.f = 0.3) # 0-1, lower values make fainter points
adjustcolor(c("red", "green"), alpha.f = 0.3) # works with named colors too!
```

If you know html colour codes, you can use them directly. For the rest of us,
`adjustcolors` saves us the hassle:

```{r scatterplot colors}
plot(Sepal.Width ~ Sepal.Length, data = iris,
     pch = c(19, 17, 15)[as.numeric(Species)],
     col = adjustcolor(c("red", "blue", "orange")[as.numeric(Species)],
                       alpha.f = 0.3))
```
 
The third way to modify points is to change their sizes. This is controlled by
the `cex` argument. By default, it is set to `1`. This is a multiplication
factor: values less than 1 will reduce points, and values greater than 1 will
increase them.

> ## Challenge
>
> How can we use the size of points in our plot to convey something interesting
> about our data? Try to add some more information to our previous plot using
> the `cex` argument.

```{r cex}
plot(Sepal.Width ~ Sepal.Length, data = iris,
     pch = 19,
     col = adjustcolor(c("red", "blue", "orange")[as.numeric(Species)],
                       alpha.f = 0.3),
     cex = Petal.Length)
```

This data set works pretty well. However, if your data includes values over 10
or so, the values will produce points that are too big to plot usefully. With a
bit of code we can tweak the size however we like. Dividing by the maximum value
will translate the range to 0-1. We can then multiply the result by a single
number to set the maximum point size.

> ## Warning
>
> If the variable you use to set cex includes 0 values, those points will be
> invisible! You can fix this by adding a small value to the scaled data. i.e.,
> cex = (0.5 + values/max(values)) * 5.

```{r cex}
plot(Sepal.Width ~ Sepal.Length, data = iris,
     pch = 19, # it's hard to compare the size of different symbols
     col = adjustcolor(c("red", "blue", "orange")[as.numeric(Species)],
                       alpha.f = 0.3),
     cex = Petal.Length/max(Petal.Length) * 100)
```

## Real Data

Real data doesn't come nicely formatted. It typically requires some processing
to get it into a consistent state that R will be able to deal with.

```{r reading real data}
library(dplyr)
its1 <- read.csv("its1.tab", stringsAsFactors = FALSE)
head(its1) ## Oops, tabs!

its1 <- read.delim("its1.tab", stringsAsFactors = FALSE)
head(its1) # better, but ID isn't data!

its1 <- read.delim("its1.tab", stringsAsFactors = FALSE,
                   row.names = 1)

meta <- read.csv("metadata.csv", row.names = 1) # strings as factors for this!
meta$Year <- factor(meta$Year, ordered = TRUE)

```

```{r cleaning our data}
otus <- select(its1, taxonomy) # collect the taxonomy column
samples <- select(its1, -taxonomy) # remove the taxonomy column
samples <- t(samples) # convert taxa x observations to observations x taxa

otus$taxonomy <- gsub(".__", "", otus$taxonomy) # gsub works on vectors

library(tidyr) # for the separate function
otus <- separate(otus, taxonomy, sep = "; ",
                 into = c("kingdom", "phylum", "class", "order",
                          "family", "genus", "species"))
```

```{r ordination}
library(vegan)
samples.nmds <- metaMDS(samples)
nmds.pts <- scores(samples.nmds)

plot(nmds.pts, col = adjustcolor(as.numeric(meta$Year), alpha.f = 0.5),
     pch = c(19, 17, 15)[as.numeric(meta$Plots)],
     cex = (8 * meta$DON / max(meta$DON)) + 1)

ef <- envfit(samples.nmds, meta, permu = 999, na.rm = TRUE)
```


## type :: what kind of symbols do you want to plot?

plot(myvect, type = "p")                # the default, points
plot(myvect, type = "l")                # a connected line
plot(myvect, type = "b")                # points and a line
plot(myvect, type = "s")                # steps
plot(myvect, type = "n")                # no symbols

## the type = "n" option is useful when you want to specify
## particular details of the symbols yourself.

## Most plot types can also take arguments for main, sub, xlab,
## and ylab.

plot(myvect, main = "My plot")
plot(myvect, main = "My plot", xlab = "Order")

## You can probably guess what ylab does. Try sub to find out
## where it goes.

## Note that each of these four arguments can be set to the empty
## string to omit the labels from your plot:

plot(myvect, xlab = "", ylab = "")

## You can also turn off all labels by setting the ann (for annotation)
## argument to false:

plot(myvect, ann = FALSE)

## Other key help pages to refer to:
?plot.default
?par


############################
## Axis limits and Aspect ##
############################

## you can set the range of each axis with xlim and ylim:

smallnums <- seq(-1, 1, length.out = 10)
bignums <- seq(-10, 10, length.out = 10)
plot(smallnums, bignums)

plot(smallnums, bignums, xlim = c(-5, 5))

## If you want to use the same scale on both axes, set asp = 1:

plot(smallnums, bignums, asp = 1)

## This is important if the absolute distances between points is
## meaningful, as in ordination plots!

###########################
## High-level Plot Types ##
###########################

hist(iris$Sepal.Length)

boxplot(iris$Sepal.Length)
boxplot(Sepal.Length ~ Species, data = iris)

stripchart(iris$Sepal.Length)
stripchart(Sepal.Length ~ Species, data = iris)
stripchart(Sepal.Length ~ Species, data = iris, method = "jitter")
stripchart(Sepal.Length ~ Species, data = iris, method = "jitter",
           vertical = TRUE)

## Dynamite plots?


#################################
## Adding Information to Plots ##
#################################

plot(Sepal.Length ~ Sepal.Width, data = iris)

#############
## Symbols ##
#############

## pch = "plot character"

plot(Sepal.Length ~ Sepal.Width, data = iris, pch = 1) # default
plot(Sepal.Length ~ Sepal.Width, data = iris, pch = 2) # custom!

## 25 different symbols available:
plot(1:25, pch = 1:25)

## You can use any ascii characters:

plot(1:26, pch = letters)

## Any numeric vector can be used to set values.
## Recall that factors are actually numbers with labels!

plot(Sepal.Length ~ Sepal.Width, data = iris, pch = as.numeric(Species))

#############
## Colours ##
#############

## col = "symbol colour"

## Eight colours can be selected by number:
plot(1:10, col = 1:10, pch = 15, cex = 4)


## Colours can also be selected by name:
plot(1:10, col = "red", pch = 15, cex = 4)
plot(1:10, col = c("blue", "red"), pch = 15, cex = 4)

## Note how the colour vector is recycled!

plot(rep(1:25, 20), rep(1:20, each = 25), pch = 15, cex = 2.5,
     col = colours(distinct = TRUE))

## colours() returns a list of all the named colours R knows

## other important colour functions:
##   palette() : change the numbered colours
##   rainbow(), heat.colors() etc : create interesting palettes
##   rgb() : create your own colours

## Putting this to work:

plot(Sepal.Length ~ Sepal.Width, data = iris, pch = as.numeric(Species),
     col = as.numeric(Species))

##########################
## cex and bubble plots ##
##########################

## cex controls the size of plot points

library(vegan)
data(dune)
data(dune.env)
dune.rda <- rda(dune)
dune.sites <- scores(dune.rda, display = "sites")
plot(dune.sites)
plot(dune.sites, cex = as.numeric(dune.env$Moisture))

#####################
## Low-level Plots ##
#####################

## All of the plots we've used so far have been 'high-level'. This means
## that they take over the entire plot window - you can't combine them
## with other plots. You can however add 'low-level' plotting commands
## to existing high-level plots. The parameters are very similar, so
## they should work fairly intuitively.

## Start with an empty plot:
plot(Sepal.Length ~ Sepal.Width, data = iris, type = 'n')

## add points for one versicolor:
points(Sepal.Length ~ Sepal.Width, data = subset(iris, Species ==
                                     "versicolor"))

## add points for virginica, setting different options:
points(Sepal.Length ~ Sepal.Width, data = subset(iris, Species ==
                                     "virginica"),
       pch = 5, col = 8)

points(Sepal.Length ~ Sepal.Width, data = subset(iris, Species ==
                                     "setosa"),
       pch = 3, col = 5)

## add any number of arbitrary points:

points(x = 4, y = 6.5, pch = "X", cex = 5, col = "turquoise")

## Adding lines:

plot(Wind ~ Temp, data = airquality)

## Calculate the regression
wind.temp.lm <- lm(Wind ~ Temp, data = airquality)

wind.temp.lm
summary(wind.temp.lm)

abline(wind.temp.lm, col = "red")

## Adding text
text(90, 18, "high", cex = 4)
text(65, 3, "low", cex = 4)


## Interactively exploring your data:

plot(Murder ~ Assault, data = USArrests)

identify(USArrests[, c("Assault", "Murder")])
identify(USArrests[, c("Assault", "Murder")], labels = row.names(USArrests))

plot(Murder ~ Assault, data = USArrests, type = 'n')

text(Murder ~ Assault, data = USArrests, labels = row.names(USArrests))

##############
## Workflow ##
##############

## Producing figures with code is a very different process than using a
## point-and-click program. It is tedious at first, but the benefit in
## the long-run is being able to reproduce the same plot for multiple
## datasets or different subsets. The key to this approach is to use
## your script files to generate your plots - change one thing at a time
## and re-send the script, don't re-enter data each time you send a plot
## command! 


###############
## Exercises ##
###############

## using the airquality data set:

## plot Wind agains Ozone
plot(Wind ~ Ozone, data = airquality)

## limit your plot to Wind less than 10
plot(Wind ~ Ozone, subset(airquality, Wind <= 10))
plot(Wind ~ Ozone, ylim = c(0, 10), airquality)

## scale the size of points to Temp
plot(Wind ~ Ozone, data = airquality, cex = 3 * scale(Temp))

## modify the previous plot such that the color of the points is
## different according to the Months variable
plot(Wind ~ Ozone, data = airquality, cex = 3 * scale(Temp),
     col = Month)

